Title Pending

### 2019-02-17 manuscript/manuscript.tex
updated introduction

Many psychiatric disorders involve distortions to representations of the external world and, as a likely consequence, pronounced changes to learning and decision making. As a normative framework, decision theory allows us to describe and pose quantita­tive questions about optimal choice behavior \citep{DayanDaw2008} and is, therefore, a critical tool for modeling, understanding, and predicting the psychological and neurobiological mechanisms underlying psychiatric behavior \citep{maia2011, HuysDawDayan2015}. Recent investigations employing this framework have provided novel and important insights into psychiatric symptoms and disorders, including anhedonia in major depression \citep{Rutledge2017}, mood instability in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016}, habitual behavior in obsessive compulsive and addictive disorders \citep{gillan2016}, and hallucinations in schizophrenia \citep{powers2017, corlett2018}.

The family of anxiety disorders (which includes generalized anxiety disorder, social anxiety disorder, panic disorder, and specific phobia) are characterized by excessive fear and worry about future negative outcomes, and represent the most common group of psychiatric disorders \citep{kessler2009}. Though the disorders typically differ in the types of objects or situations that induce fear or anxiety, all share a set of core disturbances to cognition and behavior. First, the anxiety disorders involve pessimistic inference, or a tendency to appraise threatening objects or situations as more likely and/or more severe than they objectively are. Clinical anxiety is also associated with second-order conditioning, or learning to fear neutral cues indirectly associated with threat cues. Finally, anxiety is associated with avoidance, or the tendency to flee or avoid threatening situations. As an example, an individual with a clinical phobia of dogs may show chronic worry about coming into contact with aggressive dogs (pessimistic inference), may develop fear towards parks or other public spaces where they may encounter dogs (second-order conditioning), and might exert greater effort to avoid contact with dogs or locations associated with dogs (avoidance).

Recent work has begun to shine some light into the mechanisms underlying these symptoms. For example, a number of studies have found that anxiety is correlated with increased learning from worse-than-predicted outcomes \citep{Harle2017, Garrett2018, Aylward2019}. Systematic biases in learning could result in pessimistic inference by virtue of negatively biasing those same cached values used to make predictions at the time of retrieval. This begs the question as to why learning from negative prediction errors is greater in anxiety, and others have suggested this may reflect prior expectations in anxiety; that pessimistic expectations bias learning from expectation-congruent information. This raises the possibility of circular arguments.

Other studies have attempted to provide novel insights into avoidance. \cite{Maia2010} offered a reinforcement learning-based account of maintained avoidance in rodent conditioning tasks, an empirical finding that previous theories of avoidance could not account for \citep{Krypotos2015}. They argue that sustained avoidance is possible insofar that a well-established avoidance response generates no reward prediction error, such that no new learning takes place, and avoidance continues. As is discussed elsewhere \citep{Moutoussis2017}, the framework cannot account for sustained avoidance when an agent is accidentally exposed to an extinguished outcome. Returning to the previous example, the framework cannot account for why dog phobia remains for an individual in spite of an unplanned exposure to a friend's benign dog. More broadly, the framework cannot for why avoidance, and anxiety in general, is so persistent even with exposure therapy \citep{Craske2014}.

In the remainder of this article, we present a simple decision theoretic account of how the core symptoms of anxiety may arise and be maintained. Specifically, we demonstrate that how a loss of confidence in the reliability of the self or the environment can result in the catastrophic spread of negative value in sequential decision environments. We show how, under these assumptions, pessimistic inference, second-order conditioning, and chronic avoidance naturally follow, as well as a number of other behaviors previously found to be correlated with anxiety.

>>>>

Central to the experience and maintenance of anxiety disorders are the generalization of fear and persistent avoidance. In fear generalization, the set of perceived threats expands to include secondary cues or situations associated with the primary danger. For example, a dog phobic individual may first develop a fear of dog parks, then parks in general, and finally any public space where a dog may be plausibly encountered. As a consequence, avoidance behavior becomes more prolific often at considerable cost to the anxious individual, either in foregone rewards or in effort expenditure. The same dog phobic individual may take lengthy detours to avoid proximity to feared locations and may even abandon social obligations for fear of encountering a dog. Importantly, anxious avoidance is persistent despite the many safety precautions often in place that limit or negate the possibility of danger (e.g. fences, leash laws). These symptoms are puzzling then insofar the objective risk of danger is often so low and the cost of anxious behavior so high (as is often acknowledged by the patients themselves).

From the perspective of decision theory, present choice (e.g. to approach or avoid) is contingent on assumptions about the future, such as the reliability of the environment or the set of available future choices. A common finding in clinical research is that anxiety is correlated with pessimistic expectations about the future, such that anxious individuals are more likely to endorse a belief in a lack of control over future danger. As has been noted elsewhere \cite{Krypotos2015}, a great many studies of anxious decision making have focused on avoidance behavior in response to immediate danger rather than on early avoidance of potential future threat. The latter are arguably more important to the understanding and treatment of anxiety, as much of its disruption to everyday functioning stems from repeated avoidance of anticipated threat.

Here, we explore these issues by offering a decision theoretic account of sequential choice under pessimistic future expectations. We consider optimal behavior in deterministic, finite Markov decision processes (MDP), in which agents learn reward-maximizing (or loss-minimizing) policies in grid-world environments requiring series of choices. Specifically, our model treats pessimistic expectations as the aberrant assumption that an agent will be unable to act consistently with its preferences in the future. In other words, our model assumes that agents are reward-maximizing manner in present choice but expect that such choices will be unavailable to them in the future. We show that this minor deviation from standard temporal difference learning models captures many features of anxious behavior, including exaggerated threat appraisal, fear generalization, and persistent avoidance.

In the remaining sections, we also demonstrate that this simple model can account for the findings reported in the handful of empirical studies measuring anxious choice in sequential choice environments. We then offer a speculative prediction from the model that, consistent with negative future expectations, anxious individuals should not exhibit the free choice preference observed in non-anxious individuals. Finally, we discuss the relationship of the present model to clinical theories postulating a role for control and helplessness in anxiety disorders.
